---
title: "Introduction to data table Package"
author: "Angela Kang"
date: "December 24, 2017"
output: pdf_document
---

```{r echo = FALSE, message = FALSE}
library(data.table)
```

# Chapter 1

## Introduction

A data.table is a data.frame which is a set of columns that can be of different type but must be of the same length. The goal of data.table are:

\begin{enumerate}
  \item Reduce programming time (fewer function calls, less variable name repetition)
  \item Reduce compute time (fast aggregation, update by reference)
\end{enumerate}

data.table enables ordered joins which are useful for time series. 

*Aside*: there are three main things to remember about types in `R`,

```{r}
# 1. numeric vs. integer
typeof(1)
typeof(1L)

# 2. logical vs integer
typeof(NA)
typeof(NA_integer_)

# 3. Character columns are preferred due to R's global cache
```

The general form of a data.table call is below,

```{r eval = FALSE}
DT[i, j, by]
```

The above can be read as: "Take DT, subset rows using *i*, then calculate *j* grouped by *by*"
The i is analogous to WHERE, j to SELECT and by to BY statements in SQL.

When creating a data.table there can be some recycling - shorter columns will be repeated to match the length of the longest column as shown below. The vector for `B` is repeated twice and the logical for `D` is repeated 6 times.

```{r}
DT <- data.table(A = 1:6, B = c("a", "b", "c"), C = rnorm(6), D = TRUE)
DT
```

## Selecting Rows (i)

To select rows of DT we can set the *i* "argument".

```{r}
DT[3:5,]
```

Note that `.N` can be used in the *i* argument to designate the number of rows in the data.table.

```{r}
DT[3:(.N - 1)]
```

data.table offers the advantage over a data frame that a *j* "argument" does not need to be provided at all so the below returns the same as above.

```{r}
DT[3:5]
```

## Selecting Columns (j)

To select columns of DT we can set the *j* "argument". Notice the use of `.()` which is an alias for `list()` and they mean the same in data.tables.

```{r}
DT[,.(B,C)]
```

We can also create columns from pre-existing ones and select them.

```{r}
DT[, .(Total = sum(A), Mean = mean(C))]
```

We can actually throw anything into *j*, it does not have to return a value.

```{r fig.height = 4, fig.width = 6}
DT[, plot(A, C)]
```

Note that `plot` did not have to surrounded by the `.()` syntax since we use it as a side effect and returns the value `NULL`. We can actually have multiple expressions wrapped in curly braces.

```{r fig.height = 4, fig.width = 6}
DT[, {print(A); hist(C); NULL}]
```

## Group Results (by)

To group results the *by* "argument" can be set. Notice that groups are listed in the order they are encountered.

```{r}
DT[, .(MySum = sum(A), MyMean = mean(A)), by = .(B)]
```

The *by* "argument" can also be set using a function. The below groups by odd and even numbers.

```{r}
DT[, .(MySum = sum(C)), by = .(Grp = A%%2)]
```

Since we just have one item in *j* and *by* the `.()` syntax does not need to be used if you do not care for the variable names.

```{r}
DT[, sum(C), by = A%%2]
```

# Chapter 2

## Chaining

Suppose we had the following data.table and we would like to have the new table `ans` ordered by the column `A`. 

```{r}
DT <- data.table(A = c("c","b","a"), B = 1:6)

ans <- DT[, sum(B), by = A]
ans[order(A)]
```

Instead of performing the operation after `ans` is created, we can do it in one statement with chaining.

```{r}
ans <- DT[, sum(B), by = A[order(A)]]
```

For reverse order the following adjustment can be made.

```{r}
DT[, sum(B), by = A[-order(A)]]
```

Putting all *i*, *j*, and *by* arguments together.

```{r}
DT <- data.table(x = c(rep(2, 2), rep(1, 3)),
                 y = 6:10)
DT[2:4, z := sum(y), by = x]
DT
```

## Subset of Data

Suppose we had to compute the median of each column in a data.table. This could be quite repetitive to write in the *j* "argument". data.table package includes a variable called `.SD` or "Subset of Data" which itself is a data.table. `.SD` holds the values of all columns except the one specified in the *by* "argument". The below calls are equivalent.

```{r}
DT <- as.data.table(iris)

DT[, .(Sepal.Length = median(Sepal.Length),
       Sepal.Width = median(Sepal.Width),
       Petal.Length = median(Petal.Length),
       Petal.Width = median(Petal.Width)),
   by = Species]

DT[, lapply(.SD, median), by = Species]
```

Note that the `lapply` statement is not wrapped by the `.()` alias since `lapply` already returns a list.

`.SDcols` specifies the columns of the data.table that are included in `.SD`. Using `.SDcols` allows you to apply a function to all rows of a data.table, but only to some of the columns. Supose we only wanted the median length and width of sepals in our data.table. The call would then be adjusted.

```{r}
DT[, lapply(.SD, median), .SDcols = 1:2, by = Species]
```

## Using Operator := in j Argument

The `:=` operator updates the table by reference. Suppose we had the following data.table.

```{r}
DT <- data.table(x = c(rep(1, 3), rep(2, 2)),
                 y = 6:10)
DT
```

Then observe the following. Columns `x` and `z` are set using the reverse of `x` and `10:6` respectively. This actually changes the data.table instead of simply printing a view.

```{r}
DT[, c("x", "z") := .(rev(x), 10:6)]
DT
```

If we are only calling one column then the call can be simplified.

```{r}
DT[, a := z]
DT
```

To remove columns by reference we use `NULL`. Note that when deleting a column the entire column must be removed not just select rows. In addition, columns can be referred by number instead of column name. 

```{r}
DT[, c("y", "z") := NULL]
DT
```

If instead we had a list of variables we need to make sure to include brackets around the vector name so data.table doesn't try looking for that variable in the data.table.

```{r echo = FALSE}
DT <- data.table(x = c(rep(2, 2), rep(1, 3)),
                 y = 6:10,
                 z = 10:6, 
                 a = 10:6)
```

```{r}
MyCols = c("y", "z")

# This would return an error since no MyCols variable exists in DT
# DT[, MyCols := NULL]

# This is the correct call
DT[, (MyCols) := NULL]
DT
```

Functional `:=` is just another way to lay out multiple assignments so you can get the values along with the column names and it's a little easier to read.

```{r}
DT[, ':=' (y = 6:10, z = 1)]
DT
```

## Using set()

Suppose we need to repeated update a column of a data.table. We may be tempted to use a for loop with the `:=` operator.

```{r}
DT <- data.table(x = 1:5, y = c(1, 8, 1, 1, 1), z = c(2, 5, 4, 2, 3))

for (i in 1:5) DT[i, z:= i + 1]
```

A better alternative is to use `set()` which is a loopable low-overhead version of `:=`. Note that `set()` cannot handle grouping operations. The structure of the `set()` function is: `set(DT, index, column, value)`. Where `index` is the index used in the loop, `column` is the column(s) you wish to update in the loop, `value` is how the column(s) should be updated.

```{r eval = FALSE}
DT <- data.table(x = 1:5, y = c(1, 8, 1, 1, 1), z = c(2, 5, 4, 2, 3))

for (i in 1:5) set(DT, i, 3L, i+1)
```

The function `setnames()` can be used to set or modify columns of a data.table by reference. If we had the below data.table and we wanted to rename column `y` to be `z`, we would use the `setnames()` function. No new data.tables are created.

```{r}
DT <- data.table(x = 1:5, y = letters[1:5])
setnames(DT, "y", "z")
```

The function `setcolorder()` switches the order of columns by reference. If we wanted our new column `z` to come before `x`,

```{r}
setcolorder(DT, c("z", "x"))
```

# Chapter 3

## Indexing

When data.tables are created, an index is automatically created. So the following does not actually vector scan. They create an index automatically (by default) on `A` the first time you use column `A`.

```{r eval = FALSE}
DT[A == "a"]
DT[A %in% c("a", "c")]
```

The following will be much quicker to run since we are calling the column for the second time.

```{r eval = FALSE}
DT[A == "b"]
```

## Keys

Sometimes we would like to set a key manually on a data.table. Suppose we wanted to set the key to column `A` in `DT`, then we could simply use the function `setkey()`. This sorts the data by the column we specified and changes the table by reference - we did not have to assign the results back to the data.table `DT`.

```{r}
DT <- data.table(A = c("c","b","a"),
                 B = 1:6)
setkey(DT, A)
DT
```

It's easy to check the key of data.table.

```{r}
key(DT)
```

We can think of keys as row names. If we wanted to find all the rows with "b" as its A column we could simply use the below call.

```{r}
DT["b"]
```

Note that if we tried to do this with a data frame it would return an error.

By default, when calling by keys it would return all rows with the key value. We can specify which row we want returned by setting the `mult` argument.

```{r}
DT["b", mult = "first"]
DT["b", mult = "last"]
```

We can also return on multiple keys. Note that `NA` is returned for the key "d" since it does not exist in `DT`.

```{r}
DT[c("b","d")]
```

We can control this by setting the argument `nomatch`. If `nomatch = 0` then only matching rows are returned. `nomatch = NA` is the default.

```{r}
DT[c("b","d"), nomatch = 0]
```

We can set multiple columned keys. Suppose our new key is both column `A` and `B`. 

```{r}
DT <- data.table(A = c("c","b","a"),
                 B = c(4, 1, 6, 3, 5, 2),
                 C = 1:6)
setkey(DT, A, B)
DT
```

Then when calling for specific rows the calls look like the following. Make sure to use the `.()` notation. 

```{r}
DT[.("b",5)]
```

The second call returns an `NA` since it is not found in the data.table.

```{r}
DT[.("b",6)]
```

We can also look based on the first column of the key.

```{r}
DT[.("b")]
```

As an aside the statement `by = .EACHI` allows you to run *j* for each group in which each item in *i*. Suppose we wanted the first and last rows that had "a" as its column A.

```{r}
DT[.("a"), .SD[c(1, .N)], by = .EACHI]
```

If we wanted to print each group before returning the first and last row we could run the following.

```{r eval = FALSE}
DT[c("b","a"), {print(.SD); .SD[c(1, .N)]}, by = .EACHI]
```

## Rolling Joins

Ordered joins are useful when dealing with ordered data. This is best explained using an example. Suppose we had the data.table from the above with a two-column key. As seen from above, calling it on a value that is not existent would return `NA`.

```{r}
DT
DT[.("b",4)]
```

We can set the argument `roll` to `TRUE` and it would set it to the observed value of column of `B` that's smaller than 4 which exists in the data.frame. In this case it is 2.

```{r}
DT[.("b", 4), roll = TRUE]
```

Another option is to set `roll = "nearest"` which would see which value of `B` that exists is closest to 4. In this case it is 5.

```{r}
DT[.("b", 4), roll = "nearest"]
```

To always roll forwards we set `roll = +Inf`, and to always roll backwards we set `roll = -Inf`.

```{r}
DT[.("b", 4), roll = +Inf]
DT[.("b", 4), roll = -Inf]
```

It is also possible to set a rolling window.

```{r}
DT[.("b", 4), roll = +2]
DT[.("b", 4), roll = -2]
```

It is also possible to control the rolling behavior of ends. Recall from above if we set `roll = TRUE` we get the following behavior.

```{r}
DT
DT[.("b", 7:8), roll = TRUE]
```

We get the value 5 for the column `C` since it's the last observation that corresponds to the column `B` that's smaller 7 and 8. We can stop this rolling forward behavior by setting the argument `rollends = FALSE`.

```{r}
DT[.("b", 7:8), roll = TRUE, rollends = FALSE]
```

To carry the first observation backwards the argument can be set to `rollends = TRUE`.

```{r}
DT[.("b", 7:8), roll = TRUE, rollends = TRUE]
```